Deep recurrent neural networks for supernovae classification

### 2017-01-25 paper/sn_submission_v3/ms.tex
Early-epoch added

%\documentclass{article}
%\documentclass[nofootinbib,amsmath,amssymb,10pt,eqsecnum, twocolumn]{aastex61}
\documentclass[twocolumn]{aastex61}
\usepackage{graphicx, amsmath, amsmath, amssymb}
\usepackage[caption=false]{subfig}
%\usepackage[breaklinks, colorlinks, citecolor=blue]{hyperref}
%\usepackage[maxbibnames=4]{biblatex}
%\usepackage{color}
\usepackage{float}
\usepackage{natbib,hyperref,ifthen}
\usepackage{subfigure}
%------------ May need to comment out this -----------
%\usepackage{epstopdf}
%-----------less/greater than approx eq to--------------
\newcommand{\lta}{\mbox{\small\raisebox{-0.6ex}{$\,\stackrel
{\raisebox{-.2ex}{$\textstyle <$}}{\sim}\,$}}}
\newcommand{\gta}{\mbox{\small\raisebox{-0.6ex}{$\,\stackrel
{\raisebox{-.2ex}{$\textstyle >$}}{\sim}\,$}}}   
%-------------------------------------------------------
\newcommand{\edT}{\eta_{\Delta T}}
\newcommand{\nueff}{\nu_{\rm eff}}
\newcommand{\Tobs}{\tilde{T}_A^F}
\newcommand{\sigpix}{\sigma_{\rm pix}}
\newcommand{\sigsr}{\sigma_{\rm sr}}
\newcommand{\Tfa}{\Tobs}
\newcommand{\Tmod}{\hat{T}_A^F}
\newcommand{\Planck}{{\em Planck}}
\newcommand{\WMAP}{{\em WMAP}}
\newcommand{\un}{\, \rm}
\newcommand{\al}{\alpha}
\newcommand{\be}{\beta}
\newcommand{\ga}{\gamma}
\newcommand{\de}{\delta}
\newcommand{\s}{\sigma}
\newcommand{\pd}{\partial}
\newcommand{\brpd}{\bar{\partial}}
\newcommand{\Chr}[2]{\Gamma^{#1}{}_{\!#2}}
\newcommand{\brChr}[2]{\overline{\Gamma}^{#1}{}_{\!#2}}
\newcommand{\brg}{\bar{g}}
\newcommand{\brR}{\bar{R}}
\newcommand{\brG}{\bar{G}}
\newcommand{\brT}{\bar{T}}
\newcommand{\brLambda}{\overline\Lambda}
\newcommand{\brdel}{\overline\nabla}
\newcommand{\dd}{{\rm d}}
\newcommand{\jump}[1]{ \left[ #1 \right]^+_- }
\newcommand{\avrg}[1]{ \left\langle #1 \right\rangle^+_- } 
\newcommand{\Boxbar}{\overline{\Box}}
\newcommand{\diag}{\mbox{diag}}
\newcommand{\eps}{\varepsilon}
\newcommand{\oem}{{\cal T}}
\newcommand{\brW}{\overline{\cal W}}
\newcommand{\wave}{\square}
\newcommand{\brwave}{\overline{\square}}
\newcommand{\brkap}{{\bar\kappa}}
\newcommand{\bo}{\beta_0}
\newcommand{\binf}{\beta_{\!_\infty}}
\newcommand{\boldomega}{{\bf \omega}}
\newcommand{\appleq}{\raisebox{-3pt}{\makebox[14pt]{$\stackrel{\textstyle<}
{\sim}$}}}
\newcommand{\blockmatrix}[4]{\left( \begin{array}{c|c} #1
& #2 \\ \hline #3 & #4 \end{array} \right)}
\newcommand{\mbar}{{\bar m}}
\newcommand{\beq}{\begin{equation}}
\newcommand{\eeq}{\end{equation}}
\newcommand{\btheta}{{\bf \hat{n}}}
\newcommand{\absmax}{{$|\mbox{Max}|$}}
\renewcommand{\arraystretch}{1.6}
\newcommand{\qsubrm}[2]{{#1}_{\scriptsize{\textrm{#2}}}}
\def\ee{\end{equation}}
\def\bea{\begin{eqnarray}}
\def\eea{\end{eqnarray}}
\def\bse{\begin{subequations}}

@@

\newcommand\comment[1]{\textcolor{red}{#1}}

@@

{\em Recurrent neural networks} (RNNs) are a class of artificial neural network  that can learn about sequential data. They are commonly used for tasks such speech recognition and language translation, but have several possible applications in astronomy  and cosmology for processing temporal or spatial sequential data. RNNs have several properties which makes them suitable for sequential information. The inputs to the network are flexible, and they are able to recognise patterns with noisy data (for example the context of a word in a sentence relative to others can vary, or a time stream can contain instrument noise).

@@

 Observations are of the form in Table~\ref{tab:augment}, where any missing values are denoted by a dash. In order to impute the missing value of $i$, we use {\em data augmentation} and randomly select a value between $i_1$ or $i_3$. We make 5 random augmentations of all missing data, thereby increasing the size of the dataset fivefold. We can test the importance of this by training each augmentation separately and comparing the change in accuracy, which we find is $\sim 1\%$. Training with multiple augmentations at once gives the best performance since the network learns to ignore random-filled values.

\begin{table}[htdp]
\begin{center}
\begin{tabular}{|c|c|c|c|c|}\hline  
 Time&g&r&i&z\\\hline\hline 
$t_1$&$g_1$&$r_1$&$i_1$&$z_1$\\\hline 
$t_2$&$g_2$&$r_2$&$-$&$z_2$\\\hline 
$t_3$&$g_3$&$r_3$&$i_3$& $z_3$\\\hline 
 \end{tabular}
\end{center}
\label{tab:augment}
\caption{Data augmentation of missing observations. The missing data is replaced randomly by a value between $i_1$ or $i_3$.}
\end{table}%

@@

The goal of the classifier is to determine the supernovae type in the test set. We consider two problems, (1) to categorise two classes (type-Ia vs. non-type-Ia), and (2) to categorise three classes (supernovae types-1, -2 and -3). We denote these as  `SN1a' and  `123' respectively.

@@

In Fig.~\ref{fig:loss} we show the training and tests losses for such a network, with a dropout of $0.5$, applied to type-Ia vs. non-type-Ia classification with host galaxy photo-z information. Without dropout the training loss continues to fall and the test loss rises. For 5 randomised runs, training for 200 epochs, we obtain a classification accuracy of $94.8 \pm 0.2$\%, AUC of $0.986 \pm 0.001$ and $F_1 = 0.64 \pm 0.01$. The corresponding type-Ia purity and completeness are 87\% and 92\% respectively. A summary of results for this network can be found in Tables~\ref{tab:05_SN1a} and~\ref{tab:05_123}. The inclusion of host galaxy photo-z does not significantly improve the classifier performance.

\begin{table}[t!]
\begin{subtable}
\centering
\begin{tabular}{|c | c | c | c | c |} \hline 
AUC & Accuracy & F1 & Host z \\ \hline 
$ 0.986 \pm  0.001$ &  $  94.8 \pm 0.2\% $ & $  0.64 \pm 0.01  $ & T  \\
$ 0.981 \pm  0.002$ &  $  93.6 \pm 0.3\% $ & $  0.60 \pm 0.02  $ & F  \\ \hline
\end{tabular}
\caption{\label{tab:05_SN1a} Summary of results for  type-Ia vs. non-type-Ia classification with a training fraction of 0.5. The network is a 2-layer LSTM, each with 16 hidden units and dropout of 0.5.}
\end{subtable}%
\begin{subtable}
\centering
\begin{tabular}{|c | c | c | c | c |} \hline 
AUC & Accuracy & Host z  \\ \hline 
$ 0.971  \pm 0.003 $ &  $  90.4  \pm 0.4 \% $ & T  \\
$ 0.958 \pm  0.007 $ &  $   88.8 \pm 1.0 \% $ & F \\ \hline
\end{tabular}
\caption{\label{tab:05_123} Summary of results for three class categorisation (supernovae types-1, -2 and -3). The model is the same as in Table~\ref{tab:05_SN1a}}
\end{subtable}
\end{table}

One advantage of our approach is that light curve data can be directly inputted to a {\em pre-trained} model to give very fast evaluation ($<1$s) of supernovae type. In the lower panel of Fig.~\ref{fig:lightcurve} we input the light curve, as a function of time, of a type-Ia supernovae (excluded from training) to the pre-trained 2-layer LSTM model discussed above. The classifier (type-Ia vs. non-type-Ia) is initially unsure of classification, with a type-Ia probability of around 0.5. The probability then decreases slightly, but rapidly increases near the peak of the light curve. The classifier has high confidence the supernovae is of type-Ia at around 60 days, and the final probability is excess of $99.5\%$.  This method could therefore be useful to give early indication of supernovae type in surveys.

We also test the same model using a training fraction of 0.25 (around 5000 supernovae), closer to the lower end of the number likely to be followed up for the LSST. After 5 randomised runs and training for 200 epochs we obtain an accuracy of $93.3 \pm 0.4\%$, AUC of $0.978 \pm 0.002$ and $F_1 = 0.57 \pm 0.02$. The corresponding type-Ia purity and completeness are 85\% and 87\% respectively. The $F_1$ metric has degraded by $\sim10\%$ for a reduction in data of $50\%$.

For $5.2\%$ of the representative SPCC data, the training dataset is so small that over-fitting is more severe. Using the same 2-layer LSTM network with 16 hidden units and dropout of 0.5 we find a notable increase in the test loss after $\sim 20$ epochs, but the accuracy and other metrics remain relatively constant ($F_1$ values of 0.35 to 0.4 were obtained). The reason for this apparent discrepancy is that the accuracy, say, simply takes the maximum value of the softmax output layer. For example, a 2 class problem with output probabilities [0.6, 0.4] and target [1, 0] has the same accuracy as one with output probabilities [0.8, 0.2]. The loss in the latter case would be lower however, and represents increased confidence of the network in its predictions. We therefore reject models with severe over-fitting and an increasing cross-entropy loss, at the expense of metrics such as $F_1$, and decrease the model complexity.

For a training fraction of  $5.2\%$ we find a single-layer LSTM network, with 4 hidden units, and dropout of 0.5 satisfies this criteria. For 5 randomised runs, training for 200 epochs, we obtain a classification accuracy of $85.9 \pm 1.2$\%, AUC of $0.910 \pm 0.012$ and $F_1 = 0.31 \pm 0.03$. The corresponding type-Ia purity and completeness are 72\% and 66\% respectively.

@@

We have presented a new method for performing photometric classification of supernovae. Machine learning methodology has previously been applied to SPCC classification~\cite{Newling:2010bp, Karpenka:2012pm, Lochner:2016hbn}. These employ a two-step process, where features are first extracted by some method before classification. In our approach light curves are used directly as inputs to a recurrent neural network, which is able to learn information from the sequence of observations.

Using post-SPCC data, we obtain results which are competitive with previous approaches.  Although we have trained the network on the cross-entropy loss and not the $F_1$ score, for the same sized dataset of $\sim10^3 (10^4)$ supernovae (including host galaxy photo-z), \cite{Karpenka:2012pm} obtained $F_1$ values of 0.33 (0.58), and \cite{Newling:2010bp} values of 0.42 (0.57), compared to our  0.31 (0.65). Recurrent neural networks  therefore compare well with other methods when a larger training  set is available. The performance isn't  quite as good with a smaller training set, possibly due to the network having to learn from no prior information about (noisy) light curves. The current state-of-the-art for a small training set ($\sim10^3$ supernovae) comes from a combination of SALT2 (Spectral Adaptive Light curve Template 2) template fits and boosted decision trees~\cite{Lochner:2016hbn}. It would be interesting to check how how deep learning compares to this with a larger training set.

As well as finding competitive results for the final metrics, we have shown that it is possible to give fast, early evaluation of supernovae type using pre-trained models. This is possible since the light curve can be fed to the model directly without needing any feature extraction.

There are several possibilities for future work. One of the advantages of recurrent neural networks is that inputs are agnostic, so the impact of any additional inputs could be explored. It would be possible, for example, to even pass the raw images in each filter though a convolutional network and use those as inputs. We have considered a representative training sample, but spectroscopic follow up surveys may be biased. The performance of the network could be measured against selection bias, and the results used to inform the best follow up strategy.  Further work could also be performed to optimise the early detection probability of the network.  Finally, to improve performance in the small data regime one can use {\em transfer learning.} Here, a more complex network is pre-trained on simulations or existing data from other surveys, then the weights of the network are fine-tuned on the new, smaller dataset. The simulated SPCC data used in this work are based on the DES instrument, so it would be interesting to apply transfer learning to real DES data.

@@

>>>>

%\documentclass{article}
%\documentclass[nofootinbib,amsmath,amssymb,10pt,eqsecnum, twocolumn]{aastex61}
\documentclass[twocolumn]{aastex61}
\usepackage{graphicx, amsmath, amsmath, amssymb}
\usepackage[caption=false]{subfig}
%\usepackage[breaklinks, colorlinks, citecolor=blue]{hyperref}
%\usepackage[maxbibnames=4]{biblatex}
%\usepackage{color}
\usepackage{float}
\usepackage{natbib,hyperref,ifthen}
\usepackage{subfigure}
\usepackage{array}
%------------ May need to comment out this -----------
%\usepackage{epstopdf}
%-----------less/greater than approx eq to--------------
\newcommand{\lta}{\mbox{\small\raisebox{-0.6ex}{$\,\stackrel
{\raisebox{-.2ex}{$\textstyle <$}}{\sim}\,$}}}
\newcommand{\gta}{\mbox{\small\raisebox{-0.6ex}{$\,\stackrel
{\raisebox{-.2ex}{$\textstyle >$}}{\sim}\,$}}}   
%-------------------------------------------------------
\newcommand{\edT}{\eta_{\Delta T}}
\newcommand{\nueff}{\nu_{\rm eff}}
\newcommand{\Tobs}{\tilde{T}_A^F}
\newcommand{\sigpix}{\sigma_{\rm pix}}
\newcommand{\sigsr}{\sigma_{\rm sr}}
\newcommand{\Tfa}{\Tobs}
\newcommand{\Tmod}{\hat{T}_A^F}
\newcommand{\Planck}{{\em Planck}}
\newcommand{\WMAP}{{\em WMAP}}
\newcommand{\un}{\, \rm}
\newcommand{\al}{\alpha}
\newcommand{\be}{\beta}
\newcommand{\ga}{\gamma}
\newcommand{\de}{\delta}
\newcommand{\s}{\sigma}
\newcommand{\pd}{\partial}
\newcommand{\brpd}{\bar{\partial}}
\newcommand{\Chr}[2]{\Gamma^{#1}{}_{\!#2}}
\newcommand{\brChr}[2]{\overline{\Gamma}^{#1}{}_{\!#2}}
\newcommand{\brg}{\bar{g}}
\newcommand{\brR}{\bar{R}}
\newcommand{\brG}{\bar{G}}
\newcommand{\brT}{\bar{T}}
\newcommand{\brLambda}{\overline\Lambda}
\newcommand{\brdel}{\overline\nabla}
\newcommand{\dd}{{\rm d}}
\newcommand{\jump}[1]{ \left[ #1 \right]^+_- }
\newcommand{\avrg}[1]{ \left\langle #1 \right\rangle^+_- } 
\newcommand{\Boxbar}{\overline{\Box}}
\newcommand{\diag}{\mbox{diag}}
\newcommand{\eps}{\varepsilon}
\newcommand{\oem}{{\cal T}}
\newcommand{\brW}{\overline{\cal W}}
\newcommand{\wave}{\square}
\newcommand{\brwave}{\overline{\square}}
\newcommand{\brkap}{{\bar\kappa}}
\newcommand{\bo}{\beta_0}
\newcommand{\binf}{\beta_{\!_\infty}}
\newcommand{\boldomega}{{\bf \omega}}
\newcommand{\appleq}{\raisebox{-3pt}{\makebox[14pt]{$\stackrel{\textstyle<}
{\sim}$}}}
\newcommand{\blockmatrix}[4]{\left( \begin{array}{c|c} #1
& #2 \\ \hline #3 & #4 \end{array} \right)}
\newcommand{\mbar}{{\bar m}}
\newcommand{\beq}{\begin{equation}}
\newcommand{\eeq}{\end{equation}}
\newcommand{\btheta}{{\bf \hat{n}}}
\newcommand{\absmax}{{$|\mbox{Max}|$}}
\renewcommand{\arraystretch}{1.6}
\newcommand{\qsubrm}[2]{{#1}_{\scriptsize{\textrm{#2}}}}
\def\ee{\end{equation}}
\def\bea{\begin{eqnarray}}
\def\eea{\end{eqnarray}}
\def\bse{\begin{subequations}}

@@

\newcommand\comment[1]{\textcolor{red}{#1}}
\newcommand{\specialcell}[2][c]{\renewcommand{\arraystretch}{0.8}%
  \begin{tabular}[#1]{@{}c@{}}#2\end{tabular}}

@@

{\em Recurrent neural networks} (RNNs) are a class of artificial neural network  that can learn about sequential data (for an extremely comprehensive overview see~\cite{medsker1999recurrent}). They are commonly used for tasks such speech recognition and language translation, but have several possible applications in astronomy  and cosmology for processing temporal or spatial sequential data. RNNs have several properties which makes them suitable for sequential information. The inputs to the network are flexible, and they are able to recognise patterns with noisy data (for example the context of a word in a sentence relative to others can vary, or a time stream can contain instrument noise).

@@

Observations are of the form in Table~\ref{tab:augment}, where any missing values are denoted by a dash. In order to impute the missing value of $i$, we use {\em data augmentation} and randomly select a value between $i_1$ and $i_3$. We make 5 random augmentations of all missing data, thereby increasing the size of the dataset fivefold. We can test the importance of this by training each augmentation separately and comparing the change in accuracy, which we find is $\sim 1\%$. Training with multiple augmentations at once gives the best performance since the network learns to ignore random-filled values.

\begin{table}[htdp]
\begin{center}
\begin{tabular}{|c|c|c|c|c|}\hline  
 Time&g&r&i&z\\\hline\hline 
$t_1$&$g_1$&$r_1$&$i_1$&$z_1$\\\hline 
$t_2$&$g_2$&$r_2$&$-$&$z_2$\\\hline 
$t_3$&$g_3$&$r_3$&$i_3$& $z_3$\\\hline 
 \end{tabular}
\end{center}
\caption{\label{tab:augment} Data augmentation of missing observations. The missing data is replaced randomly by a value between $i_1$ and $i_3$.}
\end{table}%

@@

The times of the observations in the light-curve are irregularly spaced and whilst this may not be optimal for the network we find that it is better to use the data padded at the end of the sequence than to place observations at similar times in similar sequence positions. There may even be hidden connections between the clustering of observation times and supernovae type, although it is hard to test for this.

The goal of the classifier is to determine the supernovae type in the test set. We consider two problems, (1) to categorise two classes (type-Ia vs. non-type-Ia), and (2) to categorise three classes (supernovae types-1, -2 and -3). We denote these as  `SN1a' and  `123' respectively. We also attempt to categorise between type-Ia and non-type-Ia supernovae using only the first six observations with $S/N>4$ and the data taken on the night of the sixth observation as described in~\cite{Kessler:2010wk}.

@@

In Fig.~\ref{fig:loss} we show the training and tests losses for such a network, with a dropout of $0.5$, applied to type-Ia vs. non-type-Ia classification with host galaxy photo-z information. Without dropout the training loss continues to fall and the test loss rises. For 5 randomised runs, training for 200 epochs, we obtain a classification accuracy of $94.9 \pm 0.2$\%, AUC of $0.986 \pm 0.001$ and $F_1 = 0.64 \pm 0.01$. The corresponding type-Ia purity and completeness are $87.3 \pm 0.8\%$ and $91.4\pm 1.1\%$ respectively. A summary of results and comparisons can be found in Table~\ref{tab:05_SN1a}. The inclusion of host galaxy photo-$z$ marginally improves the classifier performance.

\begin{table*}[t!]
\centering
\begin{tabular}{ c  c  c  c  c  c  c  c }\hline\hline
Method & Training size & AUC & Accuracy (\%) & $F_1$ & Purity (\%) & Completeness (\%) & Host $z$ \\\hline\hline
A & 10,660 & $0.986 \pm 0.001$ & $94.7 \pm 0.2$ & $0.64 \pm 0.01$ & $87.3 \pm 0.8$ & $91.4 \pm 1.1$ & True \\
A & 10,660 & $0.981 \pm 0.001$ & $93.6 \pm 0.3$ & $0.60 \pm 0.02$ & $87.4 \pm 1.7$ & $85.4 \pm 2.6$ & False \\
A & 5,330 & $0.975 \pm 0.003$ & $92.9 \pm 0.6$ & $0.57 \pm 0.03$ & $86.6\pm 2.0$ & $83.4 \pm 3.4$ & True \\
A & 5,330 & $0.973 \pm 0.002$ & $92.3 \pm 0.4$ & $0.55 \pm 0.02$ & $86.2 \pm 2.4$ & $80.8 \pm 3.8$ & False \\
B & 1,103 & $0.910 \pm 0.012$ & $85.9 \pm 0.9$ & $0.31 \pm 0.03$ & $72.4 \pm 0.4$ & $66.1 \pm 6.0$ & True \\
B & 1,103 & $0.901 \pm 0.016$ & $84.6 \pm 1.7$ & $0.28 \pm 0.05$ & $68.2 \pm 3.4$ & $66.3 \pm 5.5$ & False \\
C & $\sim$10,660 & - & - & 0.58 & 85 & 88 & True \\
C & $\sim$10,660 & - & - & 0.51 & 82 & 85 & False \\
C & 1,045 & - & - & 0.33 & 70 & 75 & True \\
C & 1,045 & - & - & 0.29 & 67 & 71 & False \\
D & $\sim$8,000 & - & - & 0.55 & - & - & True \\
D & $\sim$2,00$\underset{~}{0}$ & - & - & 0.45 & - & - & True \\
\hline\hline
E & 10,660 & $0.974 \pm 0.001$ & $90.4 \pm 0.3$ & - & $90.6 \pm 0.7$ & $86.5 \pm 0.7$ & True\\
E & 10,660 & $0.959 \pm 0.006$ & $88.5 \pm 1.1$ & - & $87.6 \pm 1.1$ & $85.9 \pm 4.1$ & False\\
F & 1,103 & $0.868 \pm 0.015$ & $78.1 \pm 0.9$ & - & $70.8 \pm 3.4$ & $70.6 \pm 4.1$ & True\\
F & 1,103 & $0.865 \pm 0.011$ & $78.0 \pm 1.2$ & - & $66.9 \pm 3.2$ & $74.5 \pm 4.2$ & False\\\hline\hline
A & 10,660 & $0.977 \pm 0.002$ & $93.1 \pm 0.4$ & $0.58 \pm 0.01$ & $88.0 \pm 1.1$ & $82.2 \pm 2.8$ & True\\
A & 10,660 & $0.970 \pm 0.001$ & $92.0 \pm 0.3$ & $0.53 \pm 0.01$ & $86.0 \pm 0.9$ & $79.5 \pm 2.2$ & False\\
B & 1,103 & $0.902 \pm 0.014$ & $85.2 \pm 1.2$ & $0.29 \pm 0.04$ & $71.5 \pm 1.6$ & $62.8 \pm 5.6$ & True\\
B & 1,103 & $0.860 \pm 0.017$ & $81.6 \pm 1.2$ & $0.21 \pm 0.02$ & $62.6 \pm 3.0$ & $57.6 \pm 2.7$ & False\\\hline\hline
E & 10,660 & $*** \pm ***$ & $*** \pm ***$ & $*** \pm ***$ & $*** \pm ***$ & $*** \pm ***$ & True\\
E & 10,660 & $*** \pm ***$ & $*** \pm ***$ & $*** \pm ***$ & $*** \pm ***$ & $*** \pm ***$ & False\\
F & 1,103 & $*** \pm ***$ & $*** \pm ***$ & $*** \pm ***$ & $*** \pm ***$ & $*** \pm ***$ & True\\
F & 1,103 & $*** \pm ***$ & $*** \pm ***$ & $*** \pm ***$ & $*** \pm ***$ & $*** \pm ***$ & False\\\hline\hline
\end{tabular}
\caption{\label{tab:05_SN1a} (Top section) Summary of results for  type-Ia vs. non-type-Ia classification with a training fraction of 0.5, 0.25 and 0.052 with comparisons to similar methods in~\cite{Karpenka:2012pm} and~\cite{Newling:2010bp}. (Second section) Summary of results for types-I,~-II~and~-III classification. (Third section) Summary of results for SPCC early-epoch challenge. (Bottom section) Summary of the results for the SPCC early-epoch challenge when classifying between types-I~,~-II~and~-III supernovae. The models used are A) Unidirectional LSTM, [16, 16] with 0.5 dropout, B) Unidirectional LSTM, [4] with 0.5 dropout, C) \cite{Karpenka:2012pm}, D) \cite{Newling:2010bp}, E) Bidirectional LSTM, [16, 16] with 0.5 dropout, F) Bidirectional LSTM, [4] with 0.5 dropout. Errors on results are the mean and standard deviation values from 5 randomised runs.}
\end{table*}

%\begin{table}[t!]
%\begin{subtable}
%\centering
%\begin{tabular}{|c | c | c | c | c |} \hline 
%AUC & Accuracy & F1 & Host z \\ \hline 
%$ 0.986 \pm  0.001$ &  $  94.8 \pm 0.2\% $ & $  0.64 \pm 0.01  $ & T  \\
%$ 0.981 \pm  0.002$ &  $  93.6 \pm 0.3\% $ & $  0.60 \pm 0.02  $ & F  \\ \hline
%\end{tabular}
%\caption{\label{tab:05_SN1a} Summary of results for  type-Ia vs. non-type-Ia classification with a training fraction of 0.5. The network is a 2-layer LSTM, each with 16 hidden units and dropout of 0.5.}
%\end{subtable}%
%\begin{subtable}
%\centering
%\begin{tabular}{|c | c | c | c | c |} \hline 
%AUC & Accuracy & Host z  \\ \hline 
%$ 0.971  \pm 0.003 $ &  $  90.4  \pm 0.4 \% $ & T  \\
%$ 0.958 \pm  0.007 $ &  $   88.8 \pm 1.0 \% $ & F \\ \hline
%\end{tabular}
%\caption{\label{tab:05_123} Summary of results for three class categorisation (supernovae types-1, -2 and -3). The model is the same as in Table~\ref{tab:05_SN1a}}
%\end{subtable}
%\end{table}

One advantage of our approach is that light curve data can be directly input to a {\em pre-trained} model to give very fast evaluation ($<1$s) of supernovae type. In the lower panel of Fig.~\ref{fig:lightcurve} we input the light curve, as a function of time, of a type-Ia supernovae (excluded from training) to the pre-trained 2-layer LSTM model discussed above. The classifier (type-Ia vs. non-type-Ia) is initially unsure of classification, with a type-Ia probability of around 0.5. The probability then decreases slightly, but rapidly increases near the peak of the light curve. The classifier has high confidence the supernovae is of type-Ia at around 60 days, and the final probability is excess of $99.5\%$.  This method could therefore be useful to give early indication of supernovae type in surveys.

We also test the same model using a training fraction of 0.25 (around 5000 supernovae), closer to the lower end of the number likely to be followed up for the LSST. After 5 randomised runs and training for 200 epochs we obtain an accuracy of $92.9 \pm 0.6\%$, AUC of $0.975 \pm 0.003$ and $F_1 = 0.57 \pm 0.03$. The corresponding type-Ia purity and completeness are $86.6\pm2.0\%$ and $83.4\pm3.4\%$ respectively. The $F_1$ metric has degraded by $\sim10\%$ for a reduction in data of $50\%$.

For $5.2\%$ of the representative SPCC data, the training dataset is so small that over-fitting is more severe. Using the same 2-layer LSTM network with 16 hidden units and dropout of 0.5 we find a notable increase in the test loss after $\sim 20$ epochs, but the accuracy and other metrics remain relatively constant ($F_1$ values of 0.35 to 0.4 were obtained). The reason for this apparent discrepancy is that the accuracy, say, simply takes the maximum value of the softmax output layer. For example, a 2 class problem with output probabilities [0.6, 0.4] and target [1, 0] has the same accuracy as one with output probabilities [0.8, 0.2]. The loss in the latter case would be lower however, and represents increased confidence of the network in its predictions. We therefore reject models with severe over-fitting and an increasing cross-entropy loss at the expense of metrics such as $F_1$, and decrease the model complexity.

For a training fraction of  $5.2\%$ we find a single-layer LSTM network, with 4 hidden units, and dropout of 0.5 satisfies this criteria. For 5 randomised runs, training for 200 epochs, we obtain a classification accuracy of $85.9 \pm 0.9$\%, AUC of $0.910 \pm 0.012$ and $F_1 = 0.31 \pm 0.03$. The corresponding type-Ia purity and completeness are $72.4\pm0.4\%$ and $66.1\pm6.0\%$ respectively.

It is difficult to directly compare the results from the SPCC directly with this work, but using post-SPCC data, we obtain results which are competitive with previous approaches. The analyses by~\cite{Karpenka:2012pm} and~\cite{Newling:2010bp} are easier to compare. Along with~\cite{Lochner:2016hbn} these employ a two-step process, where features are first extracted by various methods before machine learning classification. The results obtained for similar sized training sets are comparable as can be seen in the top section of Table~\ref{tab:05_SN1a}. When using half the dataset to train on we get a higher $F_1$ value, $F_1=0.64$ compared to $F_1=0.58$ in~\cite{Karpenka:2012pm}. The value in~\cite{Newling:2010bp} is also similar given that the sample size is smaller. For a smaller sample training set of 5.2\% of all the data we again perform similarly to~\cite{Karpenka:2012pm} but under perform compare to~\cite{Newling:2010bp} taking into account the slightly larger sample size in the latter case.

In the middle section of Table~\ref{tab:05_SN1a} the three class categorisation is shown. There is no available data for comparison of this problem, but compared to classification between type-Ia vs. non-type-Ia, bidirectional recurrent neural networks do well. The AUC and accuracy remain high, still above 90\% when the host-$z$ is included using a training fraction of 0.5

Finally the bottom section of Table~\ref{tab:05_SN1a} shows the results of the early-epoch challenge from SPCC. Here only the data before the night of the sixth observation with $S/N>4$ for each light-curve can be used - a great reduction from the use of the full light-curve. We do surprisingly well in this case obtaining an accuracy of $93.1\pm0.4\%$, AUC of $0.977\pm0.002$ and an $F_1=0.58\pm 0.01$ with a training fraction of 0.5 and including host-$z$. These values are not far from those obtained using the whole light-curve and are equivalent to the full results of~\cite{Karpenka:2012pm}. The results are not as good with a training fraction of 0.052, but still comparable to our results using the whole light-curve. The network trained on the partial light-curves does better than suggested from feeding the early-epoch light-curve through a network trained on the full sequence. This is due to the later parts of the light-curve influencing the weights of the network whilst training. Training on only the initial part of the light-curve optimises the network weights such that early sequence features have more effect, resulting in better accuracy, AUC and $F_1$ values than expected.

@@

We have presented a new method for performing photometric classification of supernovae. Machine learning methodology has previously been applied to SPCC classification~\cite{Newling:2010bp, Karpenka:2012pm, Lochner:2016hbn}. Instead of performing feature extraction before classification, our approach uses the light-curves directly as inputs to a recurrent neural network, which is able to learn information from the sequence of observations.

Although we have trained the network on the cross-entropy loss and not the $F_1$ score, for the same sized dataset of $\sim10^3 (10^4)$ supernovae (including host galaxy photo-$z$), \cite{Karpenka:2012pm} obtained $F_1$ values of 0.33 (0.58), and \cite{Newling:2010bp} values of 0.42 (0.57), compared to our  0.31 (0.64). Recurrent neural networks  therefore compare well with other methods when a larger training  set is available. The performance isn't  quite as good with a smaller training set, possibly due to the network having to learn from no prior information about (noisy) light curves. The current state-of-the-art for a small training set ($\sim10^3$ supernovae) comes from a combination of SALT2 (Spectral Adaptive Light curve Template 2) template fits and boosted decision trees~\cite{Lochner:2016hbn}. It would be interesting to check how how deep learning compares to this with a larger training set.

As well as finding competitive results for the final metrics, we have shown that it is possible to give fast, early evaluation of supernovae type using pre-trained models. This is possible since the light curve can be fed to the model directly without needing any feature extraction. Moreover, we have found that training a network only on the early epoch light-curve data results in a better early-time predictor than using a network trained on entire light-curve data.

There are several possibilities for future work. One of the advantages of recurrent neural networks is that inputs are agnostic, so the impact of any additional inputs could be explored. It would be possible, for example, to even pass the raw images in each filter though a convolutional network and use those as inputs. We have considered a representative training sample, but spectroscopic follow up surveys may be biased. The performance of the network could be measured against selection bias, and the results used to inform the best follow up strategy.  Further work could also be performed to optimise the early detection probability of the network.  Finally, to improve performance in the small data regime one can use {\em transfer learning.} Here, a more complex network is pre-trained on simulations or existing data from other surveys, then the weights of the network are fine-tuned on the new, smaller dataset. The simulated SPCC data used in this work are based on the DES instrument, and we are applying transfer learning to real DES data for publication in future work.

@@

\bibitem[{Medsker \& Jain(1999)}]{medsker1999recurrent}
Medsker, L., \& Jain, L.~C. 1999, Recurrent neural networks: design and
  applications (CRC press)
